Appendix S1. Linear regression models and non-linear population dynamics

### 2025-09-29 manuscript/Appendix_S1.Rmd
Rename Appendix code files

---
title: "Appendix S1. Linear regression models and non-linear population dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs_S1.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
library(MASS)
library(ecoevor)
library(Hmisc)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression. We use ln-N after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.

@@

```{r figs17, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs17} Linear model-predicted population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$. Fitted (line) and observed (points) values are shown.", warning=FALSE, message=FALSE}
jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
# ggplot(data = df, aes(x = N2, y = Nt1)) +
#   geom_point() +
#   geom_smooth(method = "lm", formula = y ~ x) + theme_classic()
```

We can already see that the linear model does not provide a useful fit to the data points (neither does taking the inverse function help). The regression parameters can predict well the population size of the other species, but they are not informative for inference. This has been observed previously (@Certain2018), and non-linear least squares models (among others) are more often used to estimate parameters for dynamics resulting from these kinds of species interaction models (@Kloppers2013; @Muhlbauer2020; @Olivenca2021). The inability of the linear models to reproduce the interaction coefficients is more clearly demonstrated in @Certain2018. However, they do indicate that the slope of the linear model can reflect the direction of effect for the interacting species.

@@

The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intra-specific correlation is always 1).

```{r}
# explanatory power
preds = computePredictedValues(m.4.sample)
```
```{r figs20, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs20} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions. Lines represent the 95% HPDI for predictions at each time point."}
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
points(df$Nt2,pch=19,col="orange")
lines(df$n1_25,col="black")
lines(df$n1_975,col="black")
lines(df$n2_25,col="orange")
lines(df$n2_975,col="orange")
```

@@

@@

The goal of the process was to use HMSC's latent variable approach to study the results of non-linear processes in ecology using a linear model. We can infer the direction and magnitude of the species interaction. We now consider an environmental covariate that changes over time and impacts each species growth. We will use only HMSC from this point forward, as the species interactions represent a departure from what a linear or AR model can manage.

@@

```{r}
# Initial conditions
N0 <- c(10,10)
r <- c(1.7,1.7)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
E.0 <- 0.8
x <- c(0.6,0.8)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
  E[i] <- E[i-1] + rnorm(1,0,0.1)
}
```
```{r figs21, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs21} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition and a changing environment, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
# Plot simulation: ggplot
gdat <- N
gdat$time <- 1:t
dat <- melt(gdat, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic() + ylab("N")
```
```{r}
dat <- as.data.frame(log(N))
dat$time <- 1:t
df <- data.frame(dat[(2:t),-3])
Y <- as.matrix(df)
# prepare data in HMSC format
XData <- data.frame(dat[1:(t-1),-3],E[1:(t-1)],E[1:(t-1)]^2)
colnames(XData) <- c(paste0("n",1:length(r)),"E","Esq")
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

postBeta = getPostEstimate(m.5.sample, parName = "Beta")
plotBeta(m.5.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```
We can see that the Environment has an overall negative impact for Species 1 as $E$ drifts randomly towards more positive values (and thus further from Species 1 trait value). We also see that the residual species association for species 1 and 2 are negative.

@@

```{r warning=FALSE}
VP <- computeVariancePartitioning(m.5.sample,group=c(1,1,1),groupnames="Env")
plotVariancePartitioning(m.5.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
```
We evaluate whether variation partition correctly detects the relative importance of environment and species interactions by greatly increasing the effects of interspecific competition and decreasing the strength of environmental variation.

@@

postBeta = getPostEstimate(m.6.sample, parName = "Beta")
plotBeta(m.6.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```
With the reduced impact of the environment, we now see the strong negative associations between the two competing species. We also correctly see a relatively weak impact of environment for Species 1 abundance and a stronger impact for Species 2.

@@

We now use the species abundances as the response variable in the model, with $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations). We use the absolute value of the change in trait value from one time point to the next (the total amount of trait change, regardless of direction) as a fixed effect in the HMSC.

@@

The analysis indicates that we can successfully use the HMSC analysis to partition drivers of species abundances that include evolving traits. We see that change in Species 1 trait is a strong driver of abundances in Species1 and 2, and that trait change in Species 2 has a weaker impact. This aligns well with the observed simulation results.

@@

We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment. We also add a final consideration to the model, the existence of multiple sites with correlation structure in the variance of species abundances. We consider this in our simulation by adding multiple (10) sites, with low vs. high spatial covariance in abundances, depending on distance between sites. We then include spatial random effects in the HMSC statistical model, where species associations are considered as a function of distance between sites.

@@

for (i in 2:t) {
  for(z in 1:j){
    res <- disc_LV_evol(N0=N[study$site==z & study$time==i-1,],alpha=alpha,E=E[i-1],x=x[study$site==z & study$time==i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
    N[study$site==z & study$time==i,] <- res$Nt1
    r[study$site==z & study$time==i,] <- res$r
    # trait change
    d <- E[i-1] - x[study$site==z & study$time==i-1,]
    d1 <- k*d
    x[study$site==z & study$time==i,] <- E[i-1] - d1
  }
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
}
# site-level random effect
sigma <- 0
sigma.spatial <- 2
alpha.spatial <- 0.5
Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
# draw from covariance matrix
a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
for(i in 1:t){
  N$N1[study$time==i] <- N$N1[study$time==i] + a
}
a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
for(i in 1:t){
  N$N2[study$time==i] <- N$N2[study$time==i] + a
}
```
```{r figs27, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs27} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution. Values are shown at 10 sites, with site covariance in species abundances depending on spatial distance."}
# Plot simulation: ggplot
N$site <- study$site
N$time <- study$time
dat <- melt(N, id.vars=c("site","time"))
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2) + ylab("N") + theme_classic()
```
We now fit a spatial random effect, and add that to the HMSC model.

@@

We observe a much greater fraction of variation explained by the spatial random effects.

>>>>

---
title: "Appendix S1. Linear regression models and non-linear population dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: yes
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs_S1.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
bayesplot_theme_set(new = theme_default())
library(reshape2)
library(MASS)
library(ecoevor)
library(Hmisc)
library(cowplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression. We use ln(N) after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.

@@

```{r figs17, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs17} Linear model-predicted population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$. Fitted (line) and observed (points) values are shown.", warning=FALSE, message=FALSE}
jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
# ggplot(data = df, aes(x = N2, y = Nt1)) +
#   geom_point() +
#   geom_smooth(method = "lm", formula = y ~ x) + theme_classic()
```
In the concluding part, we will compare these linear model results to a Bayesian linear model.

```{r}
# prepare data for HMSC, run later
dat <- as.data.frame(log(N))
dat$time <- 1:t
df <- data.frame(dat[(2:t),-3])
Y <- as.matrix(df)
XData <- data.frame(dat[1:(t-1),-3])
colnames(XData) <- c(paste0("n",1:length(r)))
m.3a1.hmsc = Hmsc(Y = Y[,1], XData = XData, XFormula = ~n1+n2)
m.3a2.hmsc = Hmsc(Y = Y[,2], XData = XData, XFormula = ~n1+n2)
```

We can already see that the linear model provides a useful predictive fit to the data points, but the regression parameters cannot capture the complex relationship between species densities (neither does taking the inverse function help). The regression parameters can predict well the population size of the other species, but they are not informative for inference. This has been observed previously (@Certain2018), and non-linear least squares models (among others) are more often used to estimate parameters for dynamics resulting from these kinds of species interaction models (@Kloppers2013; @Muhlbauer2020; @Olivenca2021). The inability of the linear models to reproduce the interaction coefficients is more clearly demonstrated in @Certain2018. However, they do indicate that the slope of the linear model can reflect the direction of effect for the interacting species.

@@

The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM Chapter 7 tells us the intra-specific correlation is always 1).

```{r}
# explanatory power
preds = computePredictedValues(m.4.sample)
```

```{r figs20, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs20} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions. Lines represent the 95% HPDI for predictions at each time point."}
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
points(df$Nt2,pch=19,col="orange")
lines(df$n1_25,col="black")
lines(df$n1_975,col="black")
lines(df$n2_25,col="orange")
lines(df$n2_975,col="orange")
```
In the concluding part, we will compare Bayesian linear model results.

```{r}
# prepare data for HMSC, run later
m.4a1.hmsc = Hmsc(Y = Y[,1], XData = XData, XFormula = ~X1+X2)
m.4a2.hmsc = Hmsc(Y = Y[,2], XData = XData, XFormula = ~X1+X2)
```

@@

## Conclusions

Residual species associations reflect the magnitude of species interaction strength, and the model implements these random effects to correctly predict population density.

@@

## Environmental covariate + other elements
The goal of the process was to use HMSC's latent variable approach to study the results of non-linear processes in ecology using a linear model. We can infer the direction and magnitude of the species interaction. We now consider an environmental covariate that changes over time and impacts each species growth. We will use only HMSC from this point forward, as the species interactions represent a departure from what a linear or AR model can manage.

@@

```{r}
# Initial conditions
N0 <- c(10,10)
r <- c(1.7,1.7)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
E.0 <- 0.8
x <- c(0.6,0.8)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
  E[i] <- E[i-1] + rnorm(1,0,0.1)
}
```
```{r figs21, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs21} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition and a changing environment, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
# Plot simulation: ggplot
gdat <- N
gdat$time <- 1:t
dat <- melt(gdat, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic() + ylab("N")
```

## HMSC statistical model

```{r}
dat <- as.data.frame(log(N))
dat$time <- 1:t
df <- data.frame(dat[(2:t),-3])
Y <- as.matrix(df)
# prepare data in HMSC format
XData <- data.frame(dat[1:(t-1),-3],E[1:(t-1)],E[1:(t-1)]^2)
colnames(XData) <- c(paste0("n",1:length(r)),"E","Esq")
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

postBeta = getPostEstimate(m.5.sample, parName = "Beta")
plotBeta(m.5.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```

We can see that the Environment has an overall negative impact for Species 1 as $E$ drifts randomly towards more positive values (and thus further from Species 1 trait value). We also see that the residual species association for species 1 and 2 are negative.

@@

```{r warning=FALSE}
VP <- computeVariancePartitioning(m.5.sample,group=c(1,1,1),groupnames="Env")
plotVariancePartitioning(m.5.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
```

```{r}
# prepare data for HMSC, run later
m.5a1.hmsc = Hmsc(Y = Y[,1], XData = XData, XFormula = ~n1+n2+E+Esq)
m.5a2.hmsc = Hmsc(Y = Y[,2], XData = XData, XFormula = ~n1+n2+E+Esq)
```
We evaluate whether variation partition correctly detects the relative importance of environment and species interactions by greatly increasing the effects of interspecific competition and decreasing the strength of environmental variation.

@@

postBeta = getPostEstimate(m.6.sample, parName = "Beta")
plotBeta(m.6.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```

## Conclusions
With the reduced impact of the environment, we now see the strong negative associations between the two competing species. We also correctly see a relatively weak impact of environment for Species 1 abundance and a stronger impact for Species 2.

@@

## HMSC statistical model
We now use the species abundances as the response variable in the model, with $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations). We use the absolute value of the change in trait value from one time point to the next (the total amount of trait change, regardless of direction) as a fixed effect in the HMSC.

@@

## Conclusions
The analysis indicates that we can successfully use the HMSC analysis to partition drivers of species abundances that include evolving traits. We see that change in Species 1 trait is a strong driver of abundances in Species1 and 2, and that trait change in Species 2 has a weaker impact. This aligns well with the observed simulation results.

@@

## Correlated dynamics in multiple patches
We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment. We also add a final consideration to the model, the existence of multiple sites with correlation structure in the variance of species abundances. We consider this in our simulation by adding multiple (10) sites, with low vs. high spatial covariance in abundances, depending on distance between sites. We then include spatial random effects in the HMSC statistical model, where species associations are considered as a function of distance between sites.

@@

for (i in 2:t) {
  for(z in 1:j){
    res <- disc_LV_evol(N0=N[study$site==z & study$time==i-1,],alpha=alpha,E=E[i-1],x=x[study$site==z & study$time==i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
    N[study$site==z & study$time==i,] <- res$Nt1
    r[study$site==z & study$time==i,] <- res$r
    # trait change
    d <- E[i-1] - x[study$site==z & study$time==i-1,]
    d1 <- k*d
    x[study$site==z & study$time==i,] <- E[i-1] - d1
  }
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
}
# site-level random effect
sigma <- 0
sigma.spatial <- 2
alpha.spatial <- 0.5
Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
# draw from covariance matrix
a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
for(i in 1:t){
  N$N1[study$time==i] <- N$N1[study$time==i] + a
}
a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
for(i in 1:t){
  N$N2[study$time==i] <- N$N2[study$time==i] + a
}
```
```{r figs27, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs27} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution. Values are shown at 10 sites, with site covariance in species abundances depending on spatial distance."}
# Plot simulation: ggplot
N$site <- study$site
N$time <- study$time
dat <- melt(N, id.vars=c("site","time"))
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2) + ylab("N") + theme_classic()
```

## HMSC statistical model
We now fit a spatial random effect, and add that to the HMSC model.

@@

## Conclusions

We observe a much greater fraction of variation explained by the spatial random effects when the degree of spatial autocorrelation is higher, as expected.

# Summaries across all models

```{r}
### Model A, coefficient posteriors and R2
summar <- rbind(m.post.hmsc$Beta[[1]],m.post.hmsc$Beta[[2]])
mean(summar[,2])
quantile(summar[,2],probs=c(0.05,0.95))
# Evaluate model fit
preds = computePredictedValues(m.1.sample)
MF = evaluateModelFit(hM=m.1.sample, predY=preds)
MF
```
```{r}
### Model B, coefficient posteriors and R2
summar <- rbind(m2.post.hmsc$Beta[[1]],m2.post.hmsc$Beta[[2]])
# N
mean(summar[,2])
quantile(summar[,2],probs=c(0.05,0.95))
# E
mean(summar[,3])
quantile(summar[,3],probs=c(0.05,0.95))
# E^2
mean(summar[,4])
quantile(summar[,4],probs=c(0.05,0.95))
# Evaluate model fit
preds = computePredictedValues(m.2.sample)
MF = evaluateModelFit(hM=m.2.sample, predY=preds)
MF
```
```{r}
### Model C, Scenario 1, coefficient posteriors and R2
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.3a1.sample <- sampleMcmc(m.3a1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m.3a2.sample <- sampleMcmc(m.3a2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
## N1 model
m3a1.post.hmsc <- convertToCodaObject(m.3a1.sample)
# Summary statistics for main manuscript text
summar <- rbind(m3a1.post.hmsc$Beta[[1]],m3a1.post.hmsc$Beta[[2]])
# N1
mean(summar[,2])
quantile(summar[,2],probs=c(0.05,0.95))
# N2
mean(summar[,3])
quantile(summar[,3],probs=c(0.05,0.95))
# Evaluate model fit
preds = computePredictedValues(m.3a1.sample)
MF = evaluateModelFit(hM=m.3a1.sample, predY=preds)
MF

## N2 model
m3a2.post.hmsc <- convertToCodaObject(m.3a2.sample)
# Summary statistics for main manuscript text
summar <- rbind(m3a2.post.hmsc$Beta[[1]],m3a2.post.hmsc$Beta[[2]])
# N1
mean(summar[,2])
quantile(summar[,2],probs=c(0.05,0.95))
# N2
mean(summar[,3])
quantile(summar[,3],probs=c(0.05,0.95))
# Evaluate model fit
preds = computePredictedValues(m.3a2.sample)
MF = evaluateModelFit(hM=m.3a2.sample, predY=preds)
MF
```
```{r}
### Model C, Scenario 2, coefficient posteriors and R2
# sample MCMC
m.4a1.sample <- sampleMcmc(m.4a1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m.4a2.sample <- sampleMcmc(m.4a2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
# Summary properties for main manuscript text
preds = computePredictedValues(m.4a1.sample)
MF = evaluateModelFit(hM=m.4a1.sample, predY=preds)
MF
# Summary properties for main manuscript text
preds = computePredictedValues(m.4a2.sample)
MF = evaluateModelFit(hM=m.4a2.sample, predY=preds)
MF
```

```{r}
### Model D, Scenario 1, coefficient posteriors and R2
# sample MCMC
m.5a1.sample <- sampleMcmc(m.5a1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m.5a2.sample <- sampleMcmc(m.5a2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
# Summary properties for main manuscript text
preds = computePredictedValues(m.5a1.sample)
MF = evaluateModelFit(hM=m.5a1.sample, predY=preds)
MF
# Summary properties for main manuscript text
preds = computePredictedValues(m.5a2.sample)
MF = evaluateModelFit(hM=m.5a2.sample, predY=preds)
MF
```

```{r echo=FALSE}
#############################################################
# 1. Sim: One species, logistic growth 
set.seed(42)
# Simulate initial species population growth
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)

N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r = r1.0, N0 = N[i - 1], alpha = alpha.11)
}

dat <- as.data.frame(N)
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, log(N))) + geom_point(color="skyblue") + theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)") + ylim(c(0,6.5))

#+ geom_hline(yintercept = log(((r1.0 - 1)/alpha.11)), linetype = "dashed", color = "gray")
#############################################################################
############################################################################
###########################################################################
# 2. Sim: Two species, logistic growth + environmental covariate ---------
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r = r1.0, N0 = N[i - 1], alpha = alpha.11, E = E[i - 1], x = x1.0)
  E[i] <- E[i - 1] + rnorm(1, 0, 0.1)
}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p2 <- ggplot(dat, aes(time, log(N))) + geom_point(color="skyblue") + theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)") + ylim(c(0,6.5))

#############################################################################################
####################################################################################
##################################################################################
# 3. Sim: Two species, logistic growth + competition
# Initial conditions
N0 <- c(10, 10)
r <- c(1.67, 1.7)
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
alpha <- matrix(c(alpha.11, alpha.21, alpha.12, alpha.22), nrow = 2, byrow = FALSE)
# make sure the interaction matrix is correct Simulation of model for t time
# steps
t <- 40
N <- array(NA, dim = c(t, 2))
N <- as.data.frame(N)
colnames(N) <- c("N1", "N2")
N[1, ] <- N0
for (i in 2:t) {
  N[i, ] <- disc_LV_comp(r = r, N0 = N[i - 1, ], alpha = alpha)
}

N$time <- 1:t
dat <- melt(N, id.vars="time")
p3 <- ggplot(dat, aes(time, log(value),color=variable)) + geom_point() + scale_color_manual(values = c("N1" = "skyblue", "N2" = "darkgrey"))+ theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)") + ylim(c(0,6.5))

#################################################################################################
##############################################################################################
#############################################################################################
# 4. Sim: Two species, logistic growth + competition, environmental covariate ---------
# Initial conditions
N0 <- c(10,10)
r <- c(1.7,1.7)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
E.0 <- 0.8
x <- c(0.6,0.8)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
  E[i] <- E[i-1] + rnorm(1,0,0.1)
}

### Plot of time series ###
gdat <- N
gdat$time <- 1:t
dat <- melt(gdat, id.vars="time")
p4 <- ggplot2::ggplot(dat, aes(time, log(value),color=variable)) + geom_point() + scale_color_manual(values = c("N1" = "skyblue", "N2" = "darkgrey"))+ theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)")+ ylim(c(0,4.5))

#################################################################################################
##############################################################################################
#############################################################################################
# 5. Sim: Two species, logistic growth + competition, environmental covariate, evolution
# Initial conditions
N0 <- c(10,10)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
E.0 <- 0.8
x.0 <- c(0.1,0.8)
P <- 1
w <- 2
Wmax <- 2
h2 <- 1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
}
